---
title: "Trying out MIDAS Regressions"
author: "Minna"
---

```{r, show=FALSE, output=FALSE}
library(midasr)
library(ggplot2)
# get data 
data("USrealgdp")
data("USunempr")
```

<!-- TODO: show equation of MIDAS to explain and refer to different components of the equation better-->

### Set theta/beta function
You can consturct your own theta function, which calculates how coeff should be lagged, but the midas package has pre-built theta functions too, see `?nealmon` for Normalized Exponential Almon lag MIDAS coefficients.

```{r}
# param dk = data points, aka amount of X
theta_h0 <- function(p, dk, ...) {
   i <- (1:dk-1)/100
   # pol = polynomial
   pol <- p[3]*i + p[4]*i^2
   (p[1] + p[2]*i)*exp(pol)
}

##Generate coefficients
theta0 <- theta_h0(c(-0.1,10,-10,-10),4*12)
```

The parameter function (theta0) addresses parameter proliferation by determining how to weight lags

```{r}
plot(theta0)
```

<!-- TODO: what is index 10, why index 10? -->
We see here, the weights attributed to the different X's is exponential in decay, after index 10 

### Generate X and y
```{r}
## Generate the predictor variable from arima
xx <- ts(arima.sim(model = list(ar = 0.6), 600 * 12), frequency = 12) # 600 data points
## given predictor variable and the coefficients simulate MIDAS regression response variable
y <- midas_sim(500, xx, theta0) 

x <- window(xx, start=start(y)) 

head(x,n=20)
```

### Fit MIDAS (r = restricted) model 

```{r}
# fmls here is the Beta(L, theta) function -> like hidden state
mr <- midas_r(y~fmls(x,4*12-1,12,theta_h0)-1,
              list(y=y,x=x),
              start=list(x=c(-0.1,10,-10,-10)))

##Include trend in regression -> why included with y?
mr_it <- midas_r(y~fmls(x,4*12-1,12,theta_h0)+trend,
                 list(data.frame(y=y,trend=1:500),x=x),
                 start=list(x=c(-0.1,10,-10,-10)))

summary(mr)
summary(mr_it)
```


### Forecast
```{r}
# generate new data
h <- 5
T <- length(y) # last trend val
trend_new <- (T + 1):(T + h)
x_new <- ts(arima.sim(model = list(ar = 0.6), h * 12), frequency = 12)

# forecast
mr_forecast <- midasr::forecast(mr, newdata=list(x=x_new))
mr_it_forecast <- midasr::forecast(mr_it, newdata=list(x=x_new, trend=trend_new))
# mr_forecast$mean
# mr_it_forecast$mean

summary(mr_forecast)
```


### Plot 

```{r}
df_plot <- data.frame(
  time = (T - 19):(T + h),
  actual = c(tail(y, 20), rep(NA, h)),
  mr_forecast = c(rep(NA, 20), as.numeric(mr_forecast$mean)),
  mr_it_forecast = c(rep(NA, 20), as.numeric(mr_it_forecast$mean))
)
# df_plot
```

```{r, warning=FALSE}
ggplot(df_plot, aes(x = time)) +
  geom_line(aes(y = actual), color = "black", size = 1, linetype = "solid") +
  geom_line(aes(y = mr_forecast, color = "Simple MIDAS"), size = 1) +
  geom_line(aes(y = mr_it_forecast, color = "Trend MIDAS"), size = 1) +
  labs(
    title = "MIDAS Forecast Comparison",
    y = "Y",
    color = "Model"
  ) +
  theme_minimal(base_size = 14) +
  geom_vline(xintercept = T, linetype = "dashed") +
  annotate("text", x = T + 0.5, y = max(df_plot$actual, na.rm = TRUE),
           label = "Forecast Start", hjust = 0, vjust = -0.5, size = 4)

```