title: "MF-VAR Pipeline Walkthrough"
format:
  html:
    toc: true
    toc-depth: 3
  ipynb: default
execute:
  eval: true
  echo: true
  message: false
  warning: false
freeze: auto
---

# Orientation

This Quarto notebook explains how the mixed-frequency Bayesian VAR (MF-VAR) project is wired together. It distils the main scripts under `R/`, clarifies the underlying econometric model, and gives runnable scaffolding to experiment with the workflow in `Draft_MFVAR.r`.

-   Quarterly targets: Swiss GDP growth, CPI inflation, EUR/CHF exchange rate (`R/setup.R`).
-   Monthly indicator: KOF Barometer fetched via `kofdata::get_time_series()` (`R/data_processing.R`).
-   Core engine: `mfbvar::estimate_mfbvar()` with a Minnesota prior and inverse-Wishart covariance.
-   Benchmarks: an AR(2) fallback estimated three different ways (`predict_ar2`).

> Set `execute.eval` to `true` in the YAML header when you want to run the chunks; it is disabled by default so you can read without triggering the full pipeline.

# Environment Setup

The helper in `R/setup.R` keeps package versions consistent through `renv`.

```{r}
find_project_root <- function() {
  current <- normalizePath(".", winslash = "/", mustWork = TRUE)
  repeat {
    if (file.exists(file.path(current, "Draft_MFVAR.r"))) return(current)
    parent <- dirname(current)
    if (identical(parent, current)) stop("Could not locate project root from mfvar_walkthrough.qmd.")
    current <- parent
  }
}

proj_root <- find_project_root()
setwd(proj_root)

source(file.path(proj_root, "R", "setup.R"))
source(file.path(proj_root, "R", "data_processing.R"))
source(file.path(proj_root, "R", "evaluation.R"))
source(file.path(proj_root, "R", "plotting.R"))

# Activate renv explicitly so chunks inherit the locked package versions.
source(file.path(proj_root, "renv", "activate.R"), local = TRUE)
load_required_packages(required_pkgs)

# Use lighter MF-VAR sampling settings when rendering the walkthrough so
# chunks finish quickly while still illustrating the full workflow.
options(
  mfvar.n_reps = 800L,
  mfvar.n_burnin = 400L,
  mfvar.n_thin = 2L
)
```

If any package is missing, run `renv::restore()` from the project root. The `required_pkgs` vector contains the tidyverse stack plus `mfbvar` and `kofdata`.

# Data Layer

## Quarterly Inputs

`read_quarterly_data()` loads `data/data_quarterly.csv`, verifies the expected columns, computes annualised growth rates, and returns a tibble of quarterly observations.

```{r}
library(dplyr)
library(readr)
qdat_raw <- read_quarterly_data(file.path(proj_root, "data"))
print(head(qdat_raw, 4))
```

Key steps: - Convert `%Y-%m` strings into `zoo::yearqtr` stamps. - Log-difference GDP and CPI, multiply by 400 for annualised rates. - Log EUR/CHF so the VAR works in stationary space.

## KOF Barometer

`fetch_kof_barometer()` tries two KOF API identifiers and centres the series. The result is a monthly `ts` object, which feeds the mixed-frequency prior. When working offline, plumb a cached `ts` into the function stub.

```{r}
baro_ts <- fetch_kof_barometer()
start(baro_ts); end(baro_ts); frequency(baro_ts)
```

## Aligning Frequencies

Two helpers keep the quarterly and monthly blocks consistent:

-   `trim_to_overlap()` truncates quarters beyond the available barometer history.
-   `window_baro()` extends the barometer window two months before the sample start so ragged-edge aggregation works.

```{r}
trimmed <- trim_to_overlap(qdat_raw, baro_ts)
qdat <- trimmed$qdat
baro_aligned <- window_baro(trimmed$baro_ts, qdat)
str(baro_aligned)
```

Finally, `build_Y()` arranges the inputs for `mfbvar::set_prior()`.

```{r}
Y <- build_Y(qdat, baro_aligned)
str(Y)
```

# Model Primer

## Mixed-Frequency VAR Structure

The MF-VAR combines quarterly targets `y_t` with the monthly KOF barometer `x_{m}` in a state-space system:

-   **Observation equation**: Quarterly aggregates are constructed from monthly latent states using an averaging matrix (parameter `aggregation = "average"`).
-   **State equation**: Latent monthly variables follow a VAR with `n_lags = 5` by default. The transition block includes the KOF barometer and each quarterly target.

This design lets monthly information update quarterly forecasts without waiting for full quarterly releases.

## Prior Specification

`estimate_mfvar_model()` wraps two critical calls:

1.  `mfbvar::set_prior()` builds the Minnesota prior (tightens own lags, shrinks cross-variable effects) and draws `n_reps = 4000` posterior samples after `n_burnin = 2000`, thinning every fourth draw.
2.  `mfbvar::estimate_mfbvar()` estimates the reduced-form VAR and the accompanying state-space parameters under an inverse-Wishart covariance prior (`variance = "iw"`).

Lag length and forecast horizon are configurable but jointly affect evaluation windows because the pipeline holds out the last quarters for testing.

# Forecast Workflow

The main script `Draft_MFVAR.r` orchestrates the end-to-end run:

1.  **Load helpers** (`R/setup.R`, `R/data_processing.R`, `R/evaluation.R`, `R/plotting.R`).
2.  **Prepare data**: read quarterly CSV, download barometer, align with `trim_to_overlap()` and `window_baro()`, then build `Y`.
3.  **Evaluation suites**: call `run_holdout_evaluation()` and `run_cross_validation()` to benchmark MF-VAR versus AR(2).
4.  **Model fit**: `estimate_mfvar_model(Y, n_lags = 5, n_fcst = 12)`.
5.  **Forecast tidying**: keep target variables, compute horizon labels, back-transform exchange-rate levels for presentation.
6.  **Persistence**: write CSVs, summary text, plots, and the fitted model RDS.

The snippet below mirrors the estimation block without touching disk.

```{r}
mod_ss <- estimate_mfvar_model(Y, n_lags = 5, n_fcst = 12, seed = 123)
fc <- predict(mod_ss, aggregate_fcst = TRUE, pred_bands = 0.8)
head(dplyr::select(fc, variable, time, median))
```

# Benchmarks and Diagnostics

## AR(2) Fallback

`predict_ar2()` cycles through three estimators (Yule-Walker, OLS, ARIMA). The function catches warnings and retries before giving up with `NA` forecasts. This protects the evaluation tables when the simple benchmark struggles with short samples.

```{r}
ar2_gdp <- predict_ar2(qdat$gdp_growth, n_ahead = 4, var_label = "gdp_growth", context = "demo")
ar2_gdp
```

## Holdout Evaluation

The holdout split reserves up to four quarters after accounting for lag burn-in. For each target:

-   Fit MF-VAR on the training slice and forecast the holdout quarters.
-   Fit AR(2) on the same training slice.
-   Compute RMSE and MAE using `safe_rmse()` and `safe_mae()`.

```{r}
holdout <- run_holdout_evaluation(qdat, baro_aligned, n_lags = 5, target_vars = target_variables, out_dir = tempdir())
holdout$table
```

## Rolling Cross-Validation

Cross-validation iterates over the last `min(8, nrow(qdat) - (n_lags + 2))` quarters, re-estimating the model each time and forecasting one step ahead. The loop saves fold-level predictions for further diagnostics.

```{r}
cv <- run_cross_validation(qdat, baro_aligned, n_lags = 5, target_vars = target_variables, out_dir = tempdir())
cv$table
```

# Visual Outputs

Plotting helpers in `R/plotting.R` take tidy data frames with `lower`, `median`, and `upper` columns.

-   `plot_target_forecasts()` compares MF-VAR intervals against AR(2) point forecasts.
-   `plot_target_forecasts_with_history()` anchors the first forecast to the last observation when needed so lines join smoothly.

The production script writes PNG files such as `output/forecast_gdp_growth.png` and `output/forecast_gdp_growth_context.png`.

# Reproducing the Full Run

End-to-end execution from the project root:

```{bash eval=FALSE}
Rscript Draft_MFVAR.r
```

Outputs include: - `output/mfvar_summary.txt` with posterior diagnostics and evaluation tables. - `output/mfvar_forecasts_full.csv` and `output/mfvar_forecasts_targets.csv`. - `output/mfvar_model_ss.rds` for re-use in downstream analysis. - Optional forecast charts saved as PNGs when forecasts exist for a target.

# Extending the Pipeline

Ideas for customisation: - Add new quarterly targets by editing `target_variables` in `R/setup.R` and ensuring the quarterly CSV exposes the matching columns. - Adjust `n_lags` and `n_fcst` jointly; remember longer lags shrink the available evaluation windows. - Swap priors by changing the `prior` or `variance` arguments inside `estimate_mfvar_model()`, keeping an eye on convergence diagnostics in the summary output. - Use the cross-validation fold store (`output/forecast_cross_validation_folds.csv`) to study time-varying performance or to calibrate alternative benchmarks.

For further theoretical grounding, see the course slides in `for_our_reference/State-Space_Models_(slides_day_2).pdf`, which cover state-space representations and filtering fundamentals.