---
title: "MIDAS Regressions using Swiss Data"
author: "Minna"
---

```{r, show=FALSE, output=FALSE}
rm(list = ls())
library(midasr)
library(ggplot2)
library(kofdata)
library(lubridate)
library(dplyr)
```

## Import Data

```{r}
# only for rendering??? 
# data_raw <- read.csv("../../data/data_quarterly.csv")
# metadata <- read.csv("../../data/metadata_quarterly.csv")
data_raw <- read.csv("submission/data/data_quarterly.csv")
metadata <- read.csv("submission/data/metadata_quarterly.csv")

metadata <- metadata[, c(1,8)]

# transform to ts object
data <- data_raw |> dplyr::select(c("date", "gdp"))
# all y data
y_ext <- ts(data$gdp, start = c(1992, 1), end =  c(2027, 4), frequency=4) 
# y - without forecasts
# because we diff later, go one step forward to make up for loss of 1 val for x
y <- window(y_ext, start = c(1992, 1), end =  c(2025, 3)) # quarterly
# monthly time series -> kof (swiss) economic barometer
baro <- get_time_series("ch.kof.barometer")$ch.kof.barometer
```

#### diff and align monthly series
```{r}
# align monthly series so that after differencing it starts at the first month
# of the first quarter in y
y_start <- start(y)                 # c(year, quarter)
y_start_year <- y_start[1]
y_start_quarter <- y_start[2]
first_month_of_quarter <- (y_start_quarter - 1) * 3 + 1
prev_month <- first_month_of_quarter - 1
if (prev_month == 0) {
  prev_month <- 12
  prev_year <- y_start_year - 1
} else {
  prev_year <- y_start_year
}

# fetch one extra month before the first quarter, then difference and realign
xx0 <- window(baro, start = c(prev_year, prev_month))
x <- diff(xx0)
# after diff, both have the same start -> suitable length for midas regression
x <- window(x, start = c(y_start_year, first_month_of_quarter))
```


#### check for stationarity
```{r}
# TODO: check for seasonality as well
acf(x)
```

significant at lag 1.

```{r}
acf(diff(y))
pacf(diff(y))
```

If y is decomposed, it's white noise

## Train/Test Split

This is just preliminary until general CV is ready.

```{r}
n <- length(y)
n_val <- ceiling(0.1*n)


y_train <- window(y, end = time(y)[n - n_val]) # end 2022.25
length(y_train) # 121

# compute end time to not have alignment issues
y_train_end_time <- time(y_train)[length(y_train)]  # e.g., 2022


y_val <- window(y, start = time(y)[n - n_val + 1])
length(y_val) # 14

x_train <- window(x, end = c(y_train_end_time,3)) # ends 2022,3 -> needs to be length(x)/m=length(y)
length(x_train) # 363

x_val <- window(x, start = c(y_train_end_time,5))
length(x_val) # 42

trend_train <- 1:length(y_train)
trend_val <- (length(y_train)+1) : (length(y_train)+length(y_val))
length(trend_val) # 14
```

## Fit Models

#### Midas w/ AR(1) and trend 
```{r midas}
# check length of fmls matrix 

m <- 3 # needs to be length(x)/m=length(y) so, m = 3

f <- fmls(x_train, k=2, m=3, nealmon)
length(f)

# with trend 
m_ar1_t <- midas_r(
  y_train ~ 
  trend_train + # incl. trend 
  mls(y_train, k=1, m=1) + # AR(1)
  fmls(x_train, k=2, m=3), # max lag 1 year # add nealmon sometime
  start = list(x_train = rep(0, 3))
  )

summary(m_ar1_t)
```

#### Midas w/ AR(1) and no trend 

```{r}
# without trend
m_ar1 <- midas_r(
  y_train ~ 
  mls(y_train, k=1, m=1) + # AR(1)
  fmls(x=x_train, k=2, m=3), # max lag 1 year # add nealmon sometime
  start = list(x_train = rep(0, 3))
  )

summary(m_ar1)
```

#### Midas w/ AR(p) and no trend

#### determine suitable p


## Forecast 

#### Midas w/ AR(1) and trend
```{r}
h <- 4 # one year point forecasts

# here need to constrict x_val as well so that length(x_val)/m = length(y_val)
# w/ trend
m_ar1_t_forecast <- midasr::forecast(
  m_ar1_t, 
  # need to name them the same as above, if not, doesn't work
  newdata=list(x_train=x_val, trend_train=trend_val), 
  h=4, 
  method='dynamic' # dynamic, aka uses y_hat t for forecasting y hat t+1 -> because only 1 lag used for y (see model spec above)
  )

summary(m_ar1_t_forecast)
# from now on, only actual values is forecast
m_ar1_t_forecast <- m_ar1_t_forecast$mean
```

#### Midas w/ AR(1) and no trend

```{r}
# h <- 4 # one year point forecasts

# here need to constrict x_val as well so that length(x_val)/m = length(y_val)
# w/ trend
m_ar1_forecast <- midasr::forecast(
  m_ar1, 
  # need to name them the same as above, if not, doesn't work
  newdata=list(x_train=x_val, trend_train=trend_val), 
  h=4, 
  method='dynamic' # dynamic, aka uses y_hat t for forecasting y hat t+1 -> because only 1 lag used for y (see model spec above)
  )

summary(m_ar1_forecast)
# from now on, only actual values is forecast
m_ar1_forecast <- m_ar1_forecast$mean
length(m_ar1_forecast)
```



```{r}
# h <- 4 # one year point forecasts

# here need to constrict x_val as well so that length(x_val)/m = length(y_val)
# w/ trend
m_ar1_forecast <- midasr::forecast(
  m_ar1, 
  # need to name them the same as above, if not, doesn't work
  newdata=list(x_train=x_val, trend_train=trend_val), 
  h=4, 
  method='dynamic' # dynamic, aka uses y_hat t for forecasting y hat t+1 -> because only 1 lag used for y (see model spec above)
  )

summary(m_ar1_forecast)
# from now on, only actual values is forecast
m_ar1_forecast <- m_ar1_forecast$mean
length(m_ar1_forecast)
```

### create Plot DF

```{r}
# TODO: add true data
# for this, merge y_train and y_valid together into a single 'ts' object
h <- length(m_ar1_t_forecast)
df_plot <- data.frame(
  time = (T - 19):(T + h),
  actual = c(tail(y_train, 20), rep(NA, h)),
  actual_post = c(rep(NA, 20), tail(y_val, h)),
  m_ar1_t_forecast = c(rep(NA, 20), as.numeric(m_ar1_t_forecast)),
  m_ar1_forecast = c(rep(NA, 20), as.numeric(m_ar1_forecast))
)

df_plot
```

### Plot

```{r}
ggplot(df_plot, aes(x = time)) +
  geom_line(aes(y = actual), color = "black", size = 1, linetype = "solid") +
  geom_line(aes(y = actual_post), color = "black", size = 1, linetype = "dashed") +
  geom_line(aes(y = m_ar1_forecast, color = "Simple MIDAS"), size = 1) +
  geom_line(aes(y = m_ar1_t_forecast, color = "Trend MIDAS"), size = 1) +
  labs(
    title = "MIDAS Forecast Comparison",
    y = "Y",
    color = "Model"
  ) +
  theme_minimal(base_size = 14) +
  geom_vline(xintercept = T, linetype = "dashed")

```

### Evaluation

```{r}
# Get forecast period data only
eval_df <- tail(df_plot, h) |> 
  select(-c('actual')) |>
  mutate(
    # Calculate squared errors for each model
    se_trend = (actual_post - m_ar1_t_forecast)^2,
    se_simple = (actual_post - m_ar1_forecast)^2
  )

# Calculate MSE for each model
mse_results <- data.frame(
  Model = c("MIDAS with trend", "MIDAS without trend"),
  MSE = c(
    mean(eval_df$se_trend, na.rm = TRUE),
    mean(eval_df$se_simple, na.rm = TRUE)
  ),
  RMSE = c(
    sqrt(mean(eval_df$se_trend, na.rm = TRUE)),
    sqrt(mean(eval_df$se_simple, na.rm = TRUE))
  )
)

# Show both the detailed errors and summary statistics
print("Detailed errors by time period:")
print(eval_df)
print("\nSummary MSE/RMSE by model:")
print(mse_results)
```

So for now, the model with trend performs better than the model with trend.